Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-02 paper/ms.tex
fixing #62 [ci skip]

\section{Gaussian processes}

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are a class of stochastic
models parameterized by a mean function $\mu_\bvec{\theta}(\bvec{x})$ and a
covariance or ``kernel'' function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$
parameterized by the parameters $\bvec{\theta}$ and $\bvec{\alpha}$
respectively.
Under this model, the log-likelihood of observing a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \dfmtodo{examples and
CITE}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC;
\dfmtodo{CITE}) algorithm to sample from the posterior probability density.

@@

To provide some insight for this method, we will start by working through the
simplest possible example.
In this case, we'll assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This function is identical to \eq{celerite-kernel-complex} with $J=1$, $b=0$,
and $d=0$.
To demonstrate motivate this method, we will write out the full system of
equations that we must solve to apply the inverse of $K$ in order to compute
the first term of \eq{gp-likelihood}.
In matrix notation, this can be written as
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{x} &=& \bvec{y}, \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    x_1 \\ x_2 \\ x_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix} \quad.
\end{eqnarray}
Now, if we introduce the variables
\begin{eqnarray}\eqlabel{algo-first}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,x_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,x_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,x_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,x_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,x_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
The system defined by \eq{algo-first} through \eq{algo-last} can be
rewritten as a matrix equation to show the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    x_1 \\ u_1 \\ g_1 \\ x_2 \\ u_2 \\ g_2 \\ x_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
where we will follow \citet{Ambikasaran:2015} and call this the ``extended''
system and rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{x}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the original $K$ that we
started with, it is now has a sparse banded structure that can be exploited to
solve the system efficiently.
In particular, band matrix solvers are available that can perform an
LU-decomposition of banded matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can take advantage of these algorithms to solve our system
exactly because the target vector $\bvec{x}$ is a subset of the elements of
$\bvec{x}_\mathrm{ext}$.

@@

In the case of the full \celerite\ covariance function
(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
analogy to the $u_n$ and $g_n$ that we introduced in the previous section
\begin{eqnarray}\eqlabel{full-system-first}
\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,x_n + \psi_{n,j}\,h_{n-1,j}\\
h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,x_n - \psi_{n,j}\,g_{n-1,j}\\
u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,x_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,x_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
\end{eqnarray}
with the boundary conditions
\begin{eqnarray}
    g_{0,j} = 0 \quad, \quad
    h_{0,j} = 0 \quad, \quad
    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
    v_{N,j} = 0
\end{eqnarray}
for all $j$.
Using these variables and some algebra, we find that the following expression
\begin{eqnarray}\eqlabel{full-system-last}
\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
+ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
+ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
    &=& y_n
\end{eqnarray}
is equivalent to the target matrix equation
\begin{eqnarray}
K\,\bvec{x} &=& \bvec{y} \quad.
\end{eqnarray}
\eq{full-system-first} through \eq{full-system-last} define a banded matrix
equation in the ``extended'' space and, as before, this can be used to solve
for $K^{-1}\,\bvec{y}$ and $\det K$ in $\mathcal{O}(N)$ operations.
\Figure{matrix} shows a pictorial representation of the sparsity pattern of
the extended matrix $K_\mathrm{ext}$.
Given that definition of $K_\mathrm{ext}$, the corresponding extended vectors
$\bvec{x}_\mathrm{ext}$ and $\bvec{y}_\mathrm{ext}$ are defined schematically
as
\begin{eqnarray}
{\bvec{x}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    x_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & x_2 & u_{2,j} & \cdots &
    h_{N-1,j} & x_N
\end{array}\right)
\end{eqnarray}
and
\begin{eqnarray}
{\bvec{y}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    y_1 & 0 & 0 & 0 & 0 & y_2 & 0 & \cdots &
    0 & y_N
\end{array}\right) \quad.
\end{eqnarray}

After constructing the extended matrix (using a compact storage format), the
extended matrix can be factorized using a LU-decomposition\footnote{Even
though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
Cholesky solver cannot be used for increased efficiency.} routine optimized
for band matrices.
This decomposition can be used to compute the determinant of $K$, solve
$K^{-1}\,\bvec{y}$, and subsequently calculate the marginalized likelihood in
\eq{gp-likelihood}.

>>>>

\section{Gaussian processes}\sectlabel{gps}

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are a class of stochastic
models consisting of a mean function $\mu_\bvec{\theta}(\bvec{x})$ and a
covariance or ``kernel'' function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$
parameterized by the parameters $\bvec{\theta}$ and $\bvec{\alpha}$
respectively.
Under this model, the log-likelihood of observing a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \dfmtodo{examples and
CITE}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC;
\dfmtodo{CITE}) algorithm to sample from the posterior probability density.

@@

To provide some insight for this method, we will start by working through a
simple example.
In this case, we'll assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This function is identical to \eq{celerite-kernel-complex} with $J=1$, $b=0$,
and $d=0$.
To demonstrate this method, we will write out the full system of
equations that we must solve to apply the inverse of $K$ in order to compute
the first term of \eq{gp-likelihood}.
In matrix notation, this can be written as
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{z} &=& \bvec{y}, \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ z_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix}
\end{eqnarray}
where our goal is to solve for the unknown vector \bvec{z} for a given matrix
$K$ and vector \bvec{y}.
In this equation, we have assumed that the mean function is zero but a
non-zero mean could be included by replacing \bvec{y} by
$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
Now, if we introduce the variables
\begin{eqnarray}\eqlabel{algo-first}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
The system defined by \eq{algo-first} through \eq{algo-last} can be
rewritten as a matrix equation to show the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
where we will follow \citet{Ambikasaran:2015} and call this the ``extended''
system and rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the original $K$ that we
started with, it is now has a sparse banded structure that can be exploited to
solve the system efficiently.
In particular, band matrix solvers are available that can perform an
LU-decomposition of banded matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can take advantage of these algorithms to solve our system
exactly because the target vector $\bvec{z}$ is a subset of the elements of
$\bvec{z}_\mathrm{ext}$.

@@

In the case of the full \celerite\ covariance function
(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
analogy to the $u_n$ and $g_n$ that we introduced in the previous section
\begin{eqnarray}\eqlabel{full-system-first}
\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
\end{eqnarray}
with the boundary conditions
\begin{eqnarray}
    g_{0,j} = 0 \quad, \quad
    h_{0,j} = 0 \quad, \quad
    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
    v_{N,j} = 0
\end{eqnarray}
for all $j$.
Using these variables and some algebra, we find that the following expression
\begin{eqnarray}\eqlabel{full-system-last}
\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
+ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
+ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
    &=& r_{\bvec{\theta},n}
\end{eqnarray}
is equivalent to the target matrix equation
\begin{eqnarray}
K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
\end{eqnarray}
if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
\eq{full-system-first} through \eq{full-system-last} define a banded matrix
equation in the ``extended'' space and, as before, this can be used to solve
for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
operations.
\Figure{matrix} shows a pictorial representation of the sparsity pattern of
the extended matrix $K_\mathrm{ext}$.
Given that definition of $K_\mathrm{ext}$, the corresponding extended vectors
$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
as
\begin{eqnarray}
{\bvec{z}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
    h_{N-1,j} & z_N
\end{array}\right)
\end{eqnarray}
and
\begin{eqnarray}
{\bvec{r}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
    0 & \cdots & 0 & r_{\bvec{\theta},2}
\end{array}\right) \quad.
\end{eqnarray}

After constructing the extended matrix (using a compact storage format), the
extended matrix can be factorized using a LU-decomposition\footnote{Even
though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
Cholesky solver cannot be used for increased efficiency.} routine optimized
for band matrices.
This decomposition can be used to compute the determinant of $K$, solve
$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
likelihood in \eq{gp-likelihood}.
